---
title: "lab4"
output: html_document
---

```{bash, eval = FALSE}
wget -nc https://ftp.ncbi.nlm.nih.gov/genomes/all/GCF/000/001/735/GCF_000001735.4_TAIR10.1/GCF_000001735.4_TAIR10.1_genomic.fna.gz

#using gunzip to open
gunzip GCF_000001735.4_TAIR10.1_genomic.fna.gz

#GFF file with the gene annotation
wget -nc https://ftp.ncbi.nlm.nih.gov/genomes/all/GCF/000/001/735/GCF_000001735.4_TAIR10.1/GCF_000001735.4_TAIR10.1_genomic.gtf.gz

gunzip GCF_000001735.4_TAIR10.1_genomic.gtf.gz
```


```{r, eval = FALSE}
#Items to downlaod 
BiocManager::install("EnhancedVolcano")
BiocManager::install("clusterProfiler")

lapply(c(
  "docopt", "DT", "pheatmap","GenomicFeatures", "DESeq2",
  "edgeR", "systemPipeR", "systemPipeRdata", "BiocStyle", "GO.db", "dplyr",
  "tidyr", "stringr", "Rqc", "QuasR", "DT", "ape", "clusterProfiler","biomaRt","EnhancedVolcano"
), require, character.only = TRUE)

```

#q1
```{r load data, echo=TRUE, warning=FALSE}
library(systemPipeRdata)
library(DT)

# Get the path to the directory with the raw FASTQ files
fq_path <- systemPipeRdata::pathList()$fastqdir

# Load the original, clean targets file directly from the package
meta_file_path <- system.file("extdata", "param", "targetsPE.txt", package = "systemPipeRdata")
meta_data <- read.table(meta_file_path, header = TRUE)

# Join the directory and file names
meta_data$FileName1 <- gsub('./data/', fq_path, meta_data$FileName1)
meta_data$FileName2 <- gsub('./data/', fq_path, meta_data$FileName2)

DT::datatable(meta_data, options = list(pageLength = 10))
```

#q2
```{r plots,echo=TRUE, warning=FALSE, message=FALSE}
library(Rqc)
qa_report <- rqc(path = fq_path, pattern = "\\.fastq\\.gz$")

# per-cycle Q-score boxplots
rqcCycleQualityBoxPlot(qa_report[1:12]) 
rqcCycleQualityBoxPlot(qa_report[13:24]) 
rqcCycleQualityBoxPlot(qa_report[25:36])

# per-cycle base call frequency plts
rqcCycleBaseCallsLinePlot(qa_report[1:12]) 
rqcCycleBaseCallsLinePlot(qa_report[13:24]) 
rqcCycleBaseCallsLinePlot(qa_report[25:36]) + ggtitle("Base Calls: Files 25-36")
```

#q3
```{r process_reads_final, eval = FALSE}
library(QuasR)
library(DT)

if (!dir.exists("processed_reads")) {
  dir.create("processed_reads")
}

for (i in seq_along(meta_data$FileName1)) {
  current_file <- meta_data$FileName1[i]
  base_name <- tools::file_path_sans_ext(basename(current_file), compression = TRUE)
  new_filename <- paste0(base_name, "_processed.fastq.gz")
  output_file <- file.path("processed_reads", new_filename)
  QuasR::preprocessReads(filename = current_file,
                         outputFilename = output_file,
                         truncateEndBases = 3,
                         Rpattern = 'GCCCGGGTAA',
                         nBases = 1) 

  meta_data$FileName1[i] <- output_file
}

DT::datatable(meta_data)
```


#q4
```{bash terminal, eval= FALSE}
cd processed_reads
(base) T450-XXXXXXX:processed_readsgunzip *.gz
```

```{r directories, eval= FALSE}
dir.create("sam_files", showWarnings = FALSE)
dir.create("bam_files", showWarnings = FALSE)
writeLines(c("sam_files/", "bam_files/"), con = ".gitignore")
```

```{r align_reads_simple, eval= FALSE, message=TRUE, warning=TRUE}
hisat2_index <- "./hisat2_index/tair10_1_index"
for(i in seq_along(meta_data$FileName1)){
  sample_name <- meta_data$SampleName[i]
  fastq_file <- gsub(".gz$", "", meta_data$FileName1[i])
  sam_file <- file.path("sam_files", paste0(sample_name, ".sam"))
  bam_file <- file.path("bam_files", paste0(sample_name, ".bam"))
  sorted_bam_file <- file.path("bam_files", paste0(sample_name, "_sorted.bam"))
  log_file <- file.path("bam_files", paste0(sample_name, "_hisat2.log"))

  if (!file.exists(fastq_file)) {
    message("Skipping ", sample_name, ": FASTQ file not found.")
    next 
  }

  hisat2_log <- tryCatch({
    system2("hisat2", args = c("-p", "8",
                              "-x", hisat2_index,
                              "-U", fastq_file,
                              "-S", sam_file),
            stdout = TRUE, stderr = TRUE)
  }, error = function(e) paste("HISAT2 failed:", e$message))
  
  writeLines(hisat2_log, log_file)
  tryCatch({
    # 1. Convert SAM to BAM
    system2("samtools", args = c("view", "-bS", sam_file, "-o", bam_file))
    # 2. Sort the BAM file
    system2("samtools", args = c("sort", bam_file, "-o", sorted_bam_file))
    # 3. Index the sorted BAM file
    system2("samtools", args = c("index", sorted_bam_file))
  }, error = function(e) {
    message("Samtools failed for ", sample_name, ": ", e$message)
  })
}
print("complete.")
```

#q5
```{r idos code, eval=FALSE} 
hisat2_logs_dir <- "bam_files/" #changed this to my directory

# 1. Get a list of all HISAT2 log files
log_files <- list.files(hisat2_logs_dir, pattern = "*.log", full.names = TRUE) #changed this to *.log
percent_aligned <- 1:length(log_files)

for (i in seq_along(percent_aligned)) {
  percent_aligned[i] <- readLines(log_files[i])[length(readLines(log_files[i]))]
}

align_df <- data.frame(sort(meta_data$SampleName),percent_aligned)

align_df <- align_df %>% 
  mutate(percent_aligned = as.numeric(
    stringr::str_split_i(align_df$percent_aligned, "%",1))) %>% 
  rename(sample_name = sort.meta_data.SampleName.)

head(align_df)
```

```{r alignment_stats, echo=TRUE, message=FALSE, warning=FALSE}
library(dplyr)
library(stringr)
library(DT)
library(ggplot2)

log_dir <- "bam_files/"
log_files <- list.files(log_dir, pattern = "_hisat2.log", full.names = TRUE)
sample_names <- str_replace(basename(log_files), "_hisat2.log", "")
alignment_rates <- sapply(log_files, function(file) {
  # Read the file once
  log_content <- readLines(file)
  # The alignment rate is always the last line of the hisat2 summary
  summary_line <- log_content[length(log_content)]
  # Extract just the numeric percentage and convert it to a number
  as.numeric(str_extract(summary_line, "\\d+\\.\\d+"))
})

align_df <- data.frame(
  SampleName = sample_names,
  PercentAligned = alignment_rates
)

datatable(align_df,
          caption = "Overall Alignment Rate per Sample",
          options = list(pageLength = 10, autoWidth = TRUE),
          rownames = FALSE)

ggplot(align_df, aes(y = PercentAligned)) +
  geom_boxplot(fill = "green", color = "red", outlier.colour = "blue") +
  labs(title = "Box Plot of Alignment Percentages",
       y = "Percent Aligned (%)",
       x = "") +
  theme_minimal(base_size = 14) +
  # Remove the x-axis text and ticks as they are not needed for a single boxplot
  theme(axis.text.x = element_blank(),
        axis.ticks.x = element_blank())
```

#q6
```{r Install Rsubred, eval = FALSE } 
BiocManager::install("Rsubread")
```


```{r copy+paste from idos lab, eval = FALSE}
bfiles <- list.files("bam_files/", pattern = "_sorted.bam$", full.names = TRUE) 

gene_count_list <- Rsubread::featureCounts(files = bfiles, annot.ext = "GCF_000001735.4_TAIR10.1_genomic.gtf", 
                        isGTFAnnotationFile = TRUE, 
                        allowMultiOverlap = FALSE,  
                        isPairedEnd = FALSE, nthreads = 8,
                        minMQS = 10, 
                        GTF.featureType = "exon",  
                        GTF.attrType = "gene_id" 
                        )
```

```{r create_count_table, echo=TRUE, message=FALSE, warning=FALSE}
library(Rsubread)
library(DT)

bam_files <- list.files("./bam_files", pattern = "_sorted.bam$", full.names = TRUE)
gtf_file <- "GCF_000001735.4_TAIR10.1_genomic.gtf"

gene_counts <- featureCounts(
  files = bam_files,
  annot.ext = gtf_file,
  isGTFAnnotationFile = TRUE,
  isPairedEnd = FALSE, 
  GTF.featureType = "exon",    
  GTF.attrType = "gene_id",    
  nthreads = 8                
)

count_table <- gene_counts$counts
colnames(count_table) <- gsub("_sorted.bam", "", colnames(count_table))# remove the "_sorted.bam" suffix
count_table <- count_table[rowSums(count_table) > 0, ]

DT::datatable(count_table,
          caption = "Gene Read Counts per Sample (Filtered)",
          options = list(pageLength = 10, autoWidth = TRUE))
```

#q7
```{r create_deseq_objects, message=FALSE, warning=FALSE}
library(DESeq2)
filter_and_count <- function(res_obj, comparison_name, fc_threshold = 2) {
  res_filtered <- res_obj[!is.na(res_obj$padj) & !is.na(res_obj$log2FoldChange), ]
  sig_genes <- res_filtered[abs(res_filtered$log2FoldChange) >= log2(fc_threshold), ]
  up_regulated <- sum(sig_genes$log2FoldChange > 0)
  down_regulated <- sum(sig_genes$log2FoldChange < 0)
  
  return(data.frame(
    Comparison = comparison_name,
    Up_regulated = up_regulated,
    Down_regulated = down_regulated
  ))
}
coldata <- meta_data %>% 
  dplyr::select(SampleName, SampleLong, Factor) %>% 
  dplyr::mutate(SampleLong = str_split_i(SampleLong, "\\.", 1)) %>%
  dplyr::rename(condition = SampleLong) %>%
  dplyr::mutate(condition = factor(condition),
                Factor = factor(Factor),
                SampleName = factor(SampleName))

rownames(coldata) <- coldata$SampleName
coldata$type <- factor(rep("single-read", nrow(coldata)))

coldata <- coldata[match(colnames(count_table), rownames(coldata)), ]

all(rownames(coldata) == colnames(count_table))

dds2 <- DESeqDataSetFromMatrix(countData = count_table,
                               colData = coldata,
                               design = ~ Factor)

dds2_results <- DESeq(dds2)
```

```{r final_deseq2_analysis, echo=TRUE, message=FALSE, warning=FALSE}
targets_path <- system.file("extdata", "param", "targetsPE.txt", package = "systemPipeRdata")
comp <- systemPipeR::readComp(file = targets_path, format = "matrix", delim = "-")
comparison_pairs <- comp[[1]]

results_list <- list()

for (i in 1:nrow(comparison_pairs)) {
  
  group1 <- comparison_pairs[i, 1]
  group2 <- comparison_pairs[i, 2]
  
  comparison_name <- paste(group1, "vs", group2)
  
  res <- DESeq2::results(dds2_results, contrast = c("Factor", group1, group2), alpha = 0.2)
  
  filtered_results <- filter_and_count(res, comparison_name)
  
  results_list[[comparison_name]] <- filtered_results
}

de_summary_by_factor <- do.call(rbind, results_list)

plot_data <- de_summary_by_factor %>%
  tidyr::pivot_longer(cols = c(Up_regulated, Down_regulated),
                      names_to = "Regulation",
                      values_to = "Count")

ggplot(plot_data, aes(x = Comparison, y = Count, fill = Regulation)) +
  geom_bar(stat = "identity", position = "stack") +
  coord_flip() +
  scale_fill_manual(values = c("Up_regulated" = "brown", "Down_regulated" = "lightblue")) +
  labs(
    title = "Differentially Expressed Genes by Sample Factor",
    subtitle = "Fold Change >= 2, Alpha <= 0.2",
    x = "Comparison",
    y = "Number of Differentially Expressed Genes"
  ) +
  theme_minimal(base_size = 14) +
  theme(plot.title.position = "plot")
```

#q8
The samples cluster much better now because this analysis focuses only on genes that responded to the treatment. Before the analysis, included all genes which creates noise from genes that didn’t change. Filtering out unresponsive genes lets the true results show better. Becasue of this the new heatmap shows distinct expression patterns that reflect how the plant responds differently to the different treatments.